Homicide Increases in the COVID-19 Era

### 2025-02-22 Draft1.qmd
Fixed error
Updated notebook

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"

format:

 pdf: default

fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               # Only titles that start with # or ##
geometry: margin=1in
bibliography: references.bib
---

@@

```{r, eval=FALSE}
library(sf)
library(tidygeocoder)
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer)

@@

smoothScatter(
  x    = crime_la_2010_2019$LON,
  y    = crime_la_2010_2019$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2010–2019)",
  colramp = colorRampPalette(c("white", "skyblue", "blue", "darkblue"))
)

smoothScatter(
  x    = crime_la_2020_present$LON,
  y    = crime_la_2020_present$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2020–Present)",
  colramp = colorRampPalette(c("white", "pink", "red", "darkred"))
)

@@

```{r}
# Convert Date.Rptd to Date format (ignore time part)
crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Extract the day of the week
crime_la$DayOfWeek <- weekdays(crime_la$Date)
crime_la$DayOfWeek

# Count occurrences of crimes per day
crime_counts <- table(crime_la$DayOfWeek)
crime_counts

@@

zcta_no_NAs$S1903_C03_001E <- as.integer(zcta_no_NAs$S1903_C03_001E)

palette <- brewer.pal(7, "Greens")

# setting a layout for a 2x3 grid, margins adjusted to fit the plots, oma() for outer margin area.
par(mfrow = c(2, 3), mar = c(1, 1, 1, 6), oma = c(0, 0, 3, 0), cex.main = 1.5)

yrs <- 2018:2023

for (yr in yrs) {
  # Filter by year
  subset_data <- zcta_no_NAs[zcta_no_NAs$year == yr, ]

  # Divide it into bins
  bins <- cut(subset_data$S1903_C03_001E, breaks = 7, labels = FALSE)

  # Plot
  plot(st_geometry(subset_data),
       col = palette[bins],
       border = 'grey',
       axes = FALSE)
  title(yr, line = -1)

  # Add legend
  legend("right", inset = c(-0.5, 0), legend = round(quantile(subset_data$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0),
         fill = palette, title = "Median Income ($)", bty = "n", cex = 0.8, xpd = TRUE)
}
mtext("Median Household Income by Year and Zip Code in LA", outer = TRUE, cex = 1.5, font = 2)
```

@@

>>>>

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"
fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               # Only titles that start with # or ##
geometry: margin=1in
bibliography: references.bib
format:
  pdf: default
---

@@

```{r, output=FALSE}
library(sf)
library(tidygeocoder)
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer)

@@

smoothScatter(
  x    = crime_la_2010_2019$LON,
  y    = crime_la_2010_2019$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2010–2019)",
  colramp = colorRampPalette(c("white", "skyblue", "blue", "darkblue")),
  cex.main=0.8
)

smoothScatter(
  x    = crime_la_2020_present$LON,
  y    = crime_la_2020_present$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density (2020–Present)",
  colramp = colorRampPalette(c("white", "pink", "red", "darkred")),
  cex.main=0.8
)

@@

```{r}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Extract the day of the week
crime_la$DayOfWeek <- weekdays(crime_la$Date)
# crime_la$DayOfWeek

# Count occurrences of crimes per day
crime_counts <- table(crime_la$DayOfWeek)
# crime_counts

@@

zcta_no_NAs$S1903_C03_001E <- as.integer(zcta_no_NAs$S1903_C03_001E)
```

Now, using this joined dataset, create six maps, one for each year.

```{r}
# Calculate quantiles for the entire dataset (from 2018 to 2023)
quantiles <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

# Setting a layout for a 2x3 grid
par(mfrow = c(2, 3), mar = c(1, 1, 1, 5), oma = c(0, 0, 3, 0), cex.main = 1.5)

# Define color palette
palette <- brewer.pal(7, "Greens")

# Plot for each year
yrs <- 2018:2023
for (yr in yrs) {
  # Filter by year
  subset_data <- zcta_no_NAs[zcta_no_NAs$year == yr, ]

  # Divide into consistent bins based on the global quantiles
  bins <- cut(subset_data$S1903_C03_001E, breaks = quantiles, labels = FALSE, include.lowest = TRUE)

  # Plot
  plot(st_geometry(subset_data),
       col = palette[bins],
       border = 'grey',
       axes = FALSE,
       lwd = 0.5)
  title(yr, line = -1)

  # Add legend (use the global quantiles)
  legend("right", inset = c(-.75, 0), legend = quantiles, fill = palette, title = "Median Income ($)", bty = "n", cex = 0.8, xpd = TRUE)
}

# Outer title for the whole set of plots
mtext("Median Household Income by Year and Zip Code in LA", outer = TRUE, cex = 1, font = 2)

### Interpretation
The maps demonstrate that median income in certain LA zip codes (particularly those to the west) have a much higher median household income than the rest of the city, but over the last six years, the median income in most other LA zip codes has increased. However, there are some that a typical family is at or below the poverty line. We intend to investigate if there is a spatial correlation between these poorer zip codes and the incidence of crimes, particularly homicide, and whether this is affected by pandemic conditions.

@@

## Homicide
